---
title: "Data Wrangling with `dplyr`"
subtitle: "`R` Training"
institute: The World Bank Group - DataTax
date: today
author: ""
date-format: "dddd, [the] D[<sup style='font-size:65%;font-style:italic;'>th</sup>] [of] MMMM, YYYY"
output-dir: "_site/"
format:
  revealjs:
    theme: _extensions/dime-worldbank/dime/dime.scss
    # Output file
    output-file: 3_data_wrangling.html
    code-line-numbers: true
    # Table of Contents
    toc: false
    toc_float: false
    toc-depth: 1
    toc-location: left
    toc-title: "Agenda"
    # Generate a self contained file
    self-contained: true
    self-contained-math: true
    # Turn preview links within the presentation off (all links open in a new tab)
    preview-links: true
    auto-animate: true
    # Logo and footer options
    logo: "logos/WB-DEC-Impact-horizontal-RGB-high.png"
    footer: "Code available on [GitHub](https://github.com/rdoino-wb/data_analysis_DataTable)."
---

```{r set-up}
#| echo: false
#| warning: false

library(emo)
library(countdown)
library(dplyr)
library(data.table)
library(here)
library(knitr)
library(kableExtra)
library(stringr)
library(lubridate)
library(tidyr)

# Load the datasets from Module 2
panel_vat <- fread(here("Solutions", "Data", "Intermediate", "panel_vat.csv"))
panel_cit <- fread(here("Solutions", "Data", "Intermediate", "panel_cit.csv"))
dt_firms  <- fread(here("Solutions", "Data", "Intermediate", "dt_firms.csv"))

# Create a small example dataset for demonstrations
example_vat <- panel_vat %>% 
  slice_head(n = 5) %>%
  select(firm_id, declaration_date, vat_inputs, vat_outputs)
```

------------------------------------------------------------------------

## Where We Are

> *"The best thing about R is that it was written by statisticians. The worst thing about R is that it was written by statisticians."*
>
> -- Bo Cowgill, Google

. . .

### Module Roadmap `r emo::ji("world_map")`

::: {.fragment .fade-in}
1. ✅ **R Basics**: You can speak R!
2. ✅ **Import/Export**: You can get data in and out!
3. 👉 **Data Wrangling**: Making data behave
4. 🔜 **Visualization**: Making data speak
5. 🔜 **Analysis**: Making data work
:::

------------------------------------------------------------------------

## A Tax Administrator's Daily Challenge {.center}

. . .

You receive three datasets:

::: incremental
- 📊 **VAT declarations** (100,000 records)
- 💰 **CIT filings** (50,000 records)  
- 🏢 **Firm characteristics** (150,000 firms)
:::

. . .

Your boss asks: *"Which large firms in the manufacturing sector have increasing tax gaps?"*

. . .

### How do you even start? `r emo::ji("thinking")`

------------------------------------------------------------------------

## The Data Wrangling Journey

```{mermaid}
%%| fig-align: center
%%| fig-width: 1200

graph LR
    A[😱 Messy Data] --> B[🔧 Transform]
    B --> C[🎯 Clean Data]
    C --> D[📊 Analysis]
    
    style A fill:#ff6b6b,stroke:#c92a2a
    style B fill:#4dabf7,stroke:#1864ab
    style C fill:#51cf66,stroke:#2b8a3e
    style D fill:#ffd43b,stroke:#fab005
```

. . .

::: columns
::: {.column width="50%"}
### Before `dplyr` 😰
```r
# Which firms have high VAT?
high_vat <- panel_vat[
  panel_vat$vat_outputs > 40000,
  c("firm_id", "vat_outputs")
]
# Now sort... wait, how?
high_vat <- high_vat[
  order(high_vat$vat_outputs, 
        decreasing = TRUE),
]
```
:::

::: {.column width="50%"}
### After `dplyr` 😊
```r
# Clear as natural language!
high_vat <- panel_vat %>%
  filter(vat_outputs > 40000) %>%
  select(firm_id, vat_outputs) %>%
  arrange(desc(vat_outputs))
```
:::
:::

------------------------------------------------------------------------

## The Grammar of Data Manipulation {.center}

. . .

**Think of data manipulation as writing sentences:**

. . .

> "Take the **VAT data**, then **filter** for large taxpayers, then **select** key columns, then **arrange** by amount."

. . .

### Each verb is an action:

::: columns
::: {.column width="20%"}
📝 `select()`
:::
::: {.column width="80%"}
*"I want to see these columns"*
:::
:::

::: columns
::: {.column width="20%"}
🔍 `filter()`
:::
::: {.column width="80%"}
*"I want to see these rows"*
:::
:::

::: columns
::: {.column width="20%"}
➕ `mutate()`
:::
::: {.column width="80%"}
*"I want to calculate something new"*
:::
:::

::: columns
::: {.column width="20%"}
📊 `summarize()`
:::
::: {.column width="80%"}
*"I want to aggregate my data"*
:::
:::

::: columns
::: {.column width="20%"}
🔢 `arrange()`
:::
::: {.column width="80%"}
*"I want to sort my data"*
:::
:::

------------------------------------------------------------------------

## The Pipe: Your New Best Friend

Without the pipe: 🍝 (Spaghetti code!)

```{r no_pipe, echo = TRUE}
# Read from inside out... 🤯
result <- arrange(
  select(
    filter(panel_vat, vat_outputs > 30000),
    firm_id, vat_outputs
  ),
  desc(vat_outputs)
)
```

. . .

### With the pipe: 📖 (Like reading a story!)

```{r with_pipe, echo = TRUE}
# Read from top to bottom! 
result <- panel_vat %>%
  filter(vat_outputs > 30000) %>%
  select(firm_id, vat_outputs) %>%
  arrange(desc(vat_outputs))
```

. . .

::: callout-tip
The pipe `%>%` means **"and then..."** - it takes the output from the left and feeds it as the first argument to the right!
:::

------------------------------------------------------------------------

## Let's Start Simple

### Our toy dataset:

```{r show_example, echo = FALSE}
example_vat %>%
  kable(format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                font_size = 16)
```

. . .

### Question: How do we get just the firm IDs and outputs?

. . .

```{r select_demo, echo = TRUE}
example_vat %>%
  select(firm_id, vat_outputs)
```

------------------------------------------------------------------------

## Selecting Columns: Multiple Ways

::: panel-tabset

### By Name
```{r select_name, echo = TRUE}
panel_vat %>%
  select(firm_id, vat_outputs) %>%
  head(3)
```

### By Position
```{r select_position, echo = TRUE}
panel_vat %>%
  select(1, 4) %>%  # First and fourth columns
  head(3)
```

### By Pattern
```{r select_pattern, echo = TRUE}
panel_vat %>%
  select(starts_with("vat")) %>%
  head(3)
```

### By Exclusion
```{r select_exclude, echo = TRUE}
panel_vat %>%
  select(-reimbursement_date) %>%
  head(3)
```

:::

------------------------------------------------------------------------

## Your Turn! {.exercise}

`r countdown::countdown(minutes = 3, seconds = 0, top = 0)`

### Mini Exercise: Column Selection

Using `panel_vat`, create a dataset that:
1. Includes the firm ID
2. Includes all columns that contain "vat"
3. Excludes the reimbursement date

```{r mini_ex1, echo = TRUE, eval = FALSE}
# Your code here:
my_selection <- panel_vat %>%
  _______________
```

. . .

### Solution:
```{r mini_ex1_solution, echo = TRUE}
my_selection <- panel_vat %>%
  select(firm_id, contains("vat"), -reimbursement_date)

head(my_selection, 3)
```

------------------------------------------------------------------------

## Filtering: Finding the Needles in the Haystack

### Scenario: Your boss wants to audit high-value transactions

. . .

```{r filter_basic, echo = TRUE}
# Find declarations with outputs over 40,000
high_value <- panel_vat %>%
  filter(vat_outputs > 40000)

nrow(high_value)
```

. . .

### But wait! What about complex criteria?

```{r filter_complex, echo = TRUE}
# High outputs OR suspicious input-output ratio
suspicious <- panel_vat %>%
  filter(vat_outputs > 40000 | vat_inputs/vat_outputs > 0.9)

nrow(suspicious)
```

------------------------------------------------------------------------

## Filter Operators: Your Investigation Toolkit

```{r filter_operators, echo = FALSE}
tribble(
  ~Operator, ~Meaning, ~Example,
  "==", "Equal to", "industry == 'Manufacturing'",
  "!=", "Not equal to", "region != 'North'",
  ">", "Greater than", "vat_outputs > 50000",
  ">=", "Greater than or equal", "employees >= 100",
  "<", "Less than", "tax_rate < 0.05",
  "<=", "Less than or equal", "days_late <= 30",
  "%in%", "In a set", "firm_id %in% c('FIRM_001', 'FIRM_002')",
  "is.na()", "Is missing", "is.na(reimbursement_date)",
  "&", "AND", "size == 'Large' & region == 'North'",
  "|", "OR", "late_filing == TRUE | adjustments < 0"
) %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                font_size = 14)
```

------------------------------------------------------------------------

## Creating New Variables: The Power of `mutate()`

### The Problem:
> "We need to calculate the net VAT base and the implied VAT amount for each transaction"

. . .

### The Solution:
```{r mutate_basic, echo = TRUE}
vat_calculated <- panel_vat %>%
  mutate(
    net_vat_base = vat_outputs - vat_inputs,
    implied_vat = net_vat_base * 0.18,  # 18% VAT rate
    margin_ratio = vat_inputs / vat_outputs
  ) %>%
  select(firm_id, vat_outputs, vat_inputs, net_vat_base, implied_vat, margin_ratio)

head(vat_calculated, 4)
```

------------------------------------------------------------------------

## Conditional Logic: Making Smart Calculations

### Categorizing taxpayers by size:

```{r case_when, echo = TRUE}
#| code-line-numbers: "|2-9|"
taxpayer_categories <- panel_vat %>%
  mutate(
    taxpayer_size = case_when(
      vat_outputs < 10000 ~ "Micro",
      vat_outputs < 50000 ~ "Small",
      vat_outputs < 100000 ~ "Medium",
      TRUE ~ "Large"  # Everything else
    )
  ) %>%
  select(firm_id, vat_outputs, taxpayer_size)

# Let's see the distribution
table(taxpayer_categories$taxpayer_size)
```

------------------------------------------------------------------------

## Date Magic with `lubridate` `r emo::ji("calendar")`

### Dates are tricky... but not with the right tools!

```{r date_magic, echo = TRUE}
#| code-line-numbers: "|4-11|"
# Extract useful date components
vat_with_dates <- panel_vat %>%
  mutate(
    year = year(declaration_date),
    month = month(declaration_date),
    quarter = quarter(declaration_date),
    # Days between declaration and reimbursement
    processing_days = as.numeric(
      reimbursement_date - declaration_date
    ),
    # Fiscal year (if July-June)
    fiscal_year = ifelse(month >= 7, year + 1, year)
  ) %>%
  select(firm_id, declaration_date, year, quarter, processing_days, fiscal_year)

head(vat_with_dates, 5)
```

------------------------------------------------------------------------

## Aggregation: From Transactions to Insights

### Question: What's the total VAT collection by year?

```{r summarize_basic, echo = TRUE}
yearly_summary <- panel_vat %>%
  mutate(year = year(declaration_date)) %>%
  group_by(year) %>%
  summarize(
    n_declarations = n(),
    total_outputs = sum(vat_outputs),
    total_inputs = sum(vat_inputs),
    net_vat_base = sum(vat_outputs - vat_inputs),
    implied_collection = net_vat_base * 0.18
  )

yearly_summary
```

------------------------------------------------------------------------

## The Group-By Magic ✨

### Without grouping:
```{r no_group, echo = TRUE}
panel_vat %>%
  summarize(avg_outputs = mean(vat_outputs))
```

. . .

### With grouping - The real power!
```{r with_group, echo = TRUE}
panel_vat %>%
  mutate(year = year(declaration_date)) %>%
  group_by(year) %>%
  summarize(avg_outputs = mean(vat_outputs))
```

. . .

::: callout-important
`group_by()` splits your data into groups, and subsequent operations are performed **within each group**!
:::

------------------------------------------------------------------------

## A Real Analysis: Finding Tax Gaps

Let's identify potential compliance issues:

```{r tax_gaps, echo = TRUE}
#| code-line-numbers: "|3-6|8-9|11-18|20-21|"
# Comprehensive compliance analysis
compliance_analysis <- panel_vat %>%
  # Calculate key metrics
  mutate(
    net_base = vat_outputs - vat_inputs,
    input_ratio = vat_inputs / vat_outputs,
    year = year(declaration_date)
  ) %>%
  # Group by firm and year
  group_by(firm_id, year) %>%
  # Calculate firm-year statistics
  summarize(
    declarations = n(),
    avg_net_base = mean(net_base),
    avg_input_ratio = mean(input_ratio),
    volatility = sd(net_base),
    .groups = "drop"
  ) %>%
  # Flag suspicious patterns
  mutate(
    suspicious = avg_input_ratio > 0.95 | volatility > 50000
  )

# How many suspicious cases?
sum(compliance_analysis$suspicious)
```

------------------------------------------------------------------------

## Joining the Dots: Preparing for Integration

### Check which firms appear across datasets:

```{r overlap_analysis, echo = TRUE}
# Get unique firms from each dataset
vat_firms <- unique(panel_vat$firm_id)
cit_firms <- unique(panel_cit$firm_id)
char_firms <- unique(dt_firms$firm_id)

# Create a summary
overlap_summary <- tribble(
  ~Dataset, ~`Number of Firms`,
  "VAT Declarations", length(vat_firms),
  "CIT Declarations", length(cit_firms),
  "Firm Characteristics", length(char_firms),
  "In Both VAT & CIT", length(intersect(vat_firms, cit_firms)),
  "In All Three", length(Reduce(intersect, list(vat_firms, cit_firms, char_firms)))
)

overlap_summary %>%
  kable(format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE, font_size = 16)
```

------------------------------------------------------------------------

## Exercise: Build a Taxpayer Risk Score {.exercise}

`r countdown::countdown(minutes = 15, seconds = 0, top = 0)`

Create a risk scoring system using VAT data:

1. Calculate for each firm:
   - Average input-to-output ratio
   - Number of late reimbursements (reimbursement > 90 days after declaration)
   - Coefficient of variation in outputs (sd/mean)
   
2. Create risk flags:
   - High input ratio (>0.9)
   - Frequent late reimbursements (>50%)
   - High volatility (CV > 0.5)
   
3. Assign risk score: 1 point per flag

4. Show top 10 riskiest firms

```{r exercise_risk, echo = TRUE, eval = FALSE}
# Your code here:
risk_scores <- panel_vat %>%
  mutate(
    # Calculate metrics
    
  ) %>%
  group_by(____) %>%
  summarize(
    # Aggregate metrics
    
  ) %>%
  mutate(
    # Create risk flags and score
    
  ) %>%
  # Get top 10
  
```

------------------------------------------------------------------------

## Exercise Solution

```{r exercise_solution, echo = TRUE}
#| code-line-numbers: "|2-6|7|8-14|15-20|21|"
risk_scores <- panel_vat %>%
  mutate(
    input_ratio = vat_inputs / vat_outputs,
    processing_days = as.numeric(reimbursement_date - declaration_date),
    late_reimbursement = processing_days > 90
  ) %>%
  group_by(firm_id) %>%
  summarize(
    avg_input_ratio = mean(input_ratio, na.rm = TRUE),
    pct_late_reimb = mean(late_reimbursement, na.rm = TRUE) * 100,
    avg_outputs = mean(vat_outputs, na.rm = TRUE),
    sd_outputs = sd(vat_outputs, na.rm = TRUE),
    cv_outputs = sd_outputs / avg_outputs
  ) %>%
  mutate(
    flag_high_input = avg_input_ratio > 0.9,
    flag_late_reimb = pct_late_reimb > 50,
    flag_volatility = cv_outputs > 0.5,
    risk_score = flag_high_input + flag_late_reimb + flag_volatility
  ) %>%
  arrange(desc(risk_score), desc(avg_outputs)) %>%
  slice_head(n = 10)

risk_scores %>%
  select(firm_id, risk_score, avg_input_ratio, pct_late_reimb, cv_outputs) %>%
  kable(format = "html", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 14)
```

------------------------------------------------------------------------

## Common Patterns in Tax Data Analysis

::: panel-tabset

### Time Trends
```{r pattern_time, echo = TRUE}
# Monthly collection trends
panel_vat %>%
  mutate(month = floor_date(declaration_date, "month")) %>%
  group_by(month) %>%
  summarize(
    total_base = sum(vat_outputs - vat_inputs),
    n_filings = n()
  ) %>%
  slice_head(n = 6)
```

### Sector Analysis
```{r pattern_sector, echo = TRUE}
# Join with firm characteristics for sector analysis
manufacturing_firms <- dt_firms %>%
  filter(industry == "Manufacturing") %>%
  pull(firm_id)

panel_vat %>%
  filter(firm_id %in% manufacturing_firms) %>%
  summarize(
    total_vat_base = sum(vat_outputs - vat_inputs),
    avg_transaction = mean(vat_outputs)
  )
```

### Compliance Metrics
```{r pattern_compliance, echo = TRUE}
# Late filing analysis
panel_vat %>%
  mutate(
    days_to_reimb = as.numeric(reimbursement_date - declaration_date),
    late = days_to_reimb > 60
  ) %>%
  summarize(
    late_filing_rate = mean(late, na.rm = TRUE) * 100,
    avg_days = mean(days_to_reimb, na.rm = TRUE)
  )
```

:::

------------------------------------------------------------------------

## Best Practices for Tax Data `r emo::ji("star")`

### 1. **Always validate your calculations**
```{r validate, echo = TRUE}
# Check: Do input ratios make sense?
panel_vat %>%
  mutate(input_ratio = vat_inputs / vat_outputs) %>%
  summarize(
    min_ratio = min(input_ratio),
    max_ratio = max(input_ratio),
    ratios_above_1 = sum(input_ratio > 1)
  )
```

. . .

### 2. **Handle missing values explicitly**
```{r missing, echo = TRUE}
# Always use na.rm = TRUE in aggregations
panel_vat %>%
  summarize(
    with_na = mean(vat_outputs),        # Dangerous!
    without_na = mean(vat_outputs, na.rm = TRUE)  # Safe!
  )
```

------------------------------------------------------------------------

## Advanced Tips: Window Functions

### Ranking within groups:
```{r window_rank, echo = TRUE}
# Top 3 taxpayers by month
top_monthly <- panel_vat %>%
  mutate(
    month = floor_date(declaration_date, "month"),
    net_base = vat_outputs - vat_inputs
  ) %>%
  group_by(month) %>%
  mutate(rank = dense_rank(desc(net_base))) %>%
  filter(rank <= 3) %>%
  select(month, firm_id, net_base, rank) %>%
  arrange(month, rank)

head(top_monthly, 9)
```

------------------------------------------------------------------------

## Debugging Your Pipeline `r emo::ji("bug")`

### When things go wrong, debug step by step:

```{r debug_demo, echo = TRUE}
# Original attempt - something's wrong!
result <- panel_vat %>%
  filter(vat_outputs > 50000) %>%
  group_by(firm_id) %>%
  summarize(avg = mean(vat_outputs))

# Debug by checking each step:
# Step 1: How many rows after filter?
panel_vat %>%
  filter(vat_outputs > 50000) %>%
  nrow()

# Step 2: Check grouping
panel_vat %>%
  filter(vat_outputs > 50000) %>%
  group_by(firm_id) %>%
  tally() %>%
  head()
```

------------------------------------------------------------------------

## Performance Matters! `r emo::ji("rocket")`

### For large datasets (>1M rows):

::: columns
::: {.column width="50%"}
#### Option 1: `data.table`
```{r dt_example, echo = TRUE, eval = FALSE}
library(data.table)
dt <- as.data.table(panel_vat)

# data.table syntax
dt[vat_outputs > 50000, 
   .(avg = mean(vat_outputs)), 
   by = firm_id]
```
:::

::: {.column width="50%"}
#### Option 2: `dtplyr`
```{r dtplyr_example, echo = TRUE, eval = FALSE}
library(dtplyr)

# Use dplyr syntax with data.table speed!
panel_vat %>%
  lazy_dt() %>%
  filter(vat_outputs > 50000) %>%
  group_by(firm_id) %>%
  summarize(avg = mean(vat_outputs)) %>%
  as_tibble()
```
:::
:::

------------------------------------------------------------------------

## Your Data Wrangling Toolkit `r emo::ji("toolbox")`

### Core Verbs Recap:

```{r toolkit_table, echo = FALSE}
tribble(
  ~Verb, ~What_it_does, ~Real_Example,
  "select()", "Pick columns", "Get taxpayer ID and amounts",
  "filter()", "Pick rows", "Find large taxpayers",
  "mutate()", "Add columns", "Calculate tax rates",
  "summarize()", "Aggregate", "Total tax by region",
  "group_by()", "Split data", "Analyze by sector",
  "arrange()", "Sort rows", "Rank by tax paid"
) %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                font_size = 16) %>%
  column_spec(1, bold = TRUE, color = "white", background = "#4090BE") %>%
  column_spec(2, width = "200px") %>%
  column_spec(3, width = "300px")
```

------------------------------------------------------------------------

## Key Takeaways `r emo::ji("key")`

::: incremental

1. **Think in verbs**: Each `dplyr` function does ONE thing well

2. **Use the pipe**: Makes your code readable and debuggable

3. **Start simple**: Build complex analyses step by step

4. **Validate often**: Check your assumptions with summaries

5. **Group wisely**: Remember to `ungroup()` when done

:::

. . .

::: {.callout-tip}
## Pro tip
The best data wranglers aren't the ones who write complex code - they're the ones who write code that others (including future them) can understand!
:::

------------------------------------------------------------------------

## What's Next? `r emo::ji("crystal_ball")`

### Module 4: Data Visualization with `ggplot2`

We'll learn to:
- Transform your wrangled data into insights
- Create publication-ready visualizations
- Build dashboards for tax monitoring
- Tell stories with data

. . .

### But first... Practice! `r emo::ji("muscle")`

Try the additional exercises in `Exercises/data_wrangling_practice.R`

------------------------------------------------------------------------

## Final Challenge {.exercise}

`r countdown::countdown(minutes = 5, seconds = 0, top = 0)`

Using all three datasets, answer: *"What's the average effective CIT rate for large manufacturing firms?"*

Hints:
- Large = more than 100 employees
- Manufacturing firms from `dt_firms`
- Effective rate from `panel_cit`
- You'll need to identify common firms

```{r final_challenge, echo = TRUE, eval = FALSE}
# Your multi-dataset analysis here:

```

. . .

### Solution preview:
```{r final_solution, echo = TRUE}
# We'll explore joins in detail next module!
# For now, here's the approach:

# Step 1: Get large manufacturing firms
large_manufacturing <- dt_firms %>%
  filter(industry == "Manufacturing", 
         number_of_employees > 100) %>%
  pull(firm_id)

# Step 2: Calculate effective rates for these firms
panel_cit %>%
  filter(firm_id %in% large_manufacturing,
         taxable_income > 0) %>%
  mutate(effective_rate = (tax_paid / taxable_income) * 100) %>%
  summarize(
    avg_effective_rate = mean(effective_rate, na.rm = TRUE),
    n_firms = n_distinct(firm_id)
  )
```

---

::: {style="text-align: center; margin-top: 100px; font-size: 1.5em;"}
**Thank you!** `r emo::ji("clap")`

Questions? `r emo::ji("raising_hand")`

**Remember**: Great analysis starts with clean data! `r emo::ji("sparkles")`
:::